## Clusteing ditances from blastp results 
In this qmd we are testing diferent methods od clustering to evalute distances that are obatin from alignments between homolpgs sequences, we previously need a disimilarity matrix for having the information to make a clustering and then acces a dendrogram.  

First we load all the packages needed for running the clustering  

```{r}
#make all the inputs
library(cluster)
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(dendextend))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(corrplot))
#We read the previous generated distance matrix
mtrx_dis <- read.table(
  "results/H1_seqs_dist_matrix.tsv",
  header = TRUE,
  sep = "\t",
  row.names = 1
)
dist_obj <- as.dist(mtrx_dis)

head(mtrx_dis)
``` 


### Functions 

```{r}
plot_scat <- function() {}

predict_clust <- function(data, method_i, hc_met = "ward.D2") {
  plot_inf <- fviz_nbclust(
    data,
    FUN = hcut,
    hc_func = "hclust",
    hc_method = hc_met,
    method = method_i,
    k.max = 6
  ) +
    labs(subtitle = paste(method_i, "with", hc_met, "merge"))
  print(plot_inf)
}


save_tree <- function(clust_obj, save = TRUE, name = "Tree") {
  tree_plot <- plot(clust_obj, hang = -1, main = paste(name, "dendrogram"))

  #fviz_cluster(list(data = InputData, cluster = cls3))
  print(tree_plot)
  if (save) {
    #Save tree in Newick format
    my_tree <- as.phylo(tree_plot)
    #we can save it as a format we can visualize with programs like fig_tree
    write.tree(
      phy = my_tree,
      file = paste("results/", name, "_dendrogram.nwk", sep = "")
    )
  }
}
```

### Predict the clusters 

Now we can use diferent methods for try to determine the number of clusters in the data 
```{r} 

merge_methods = c("ward.D2", "complete", "single", "average")
infer_methods = c("gap_stat", "wss", "silhouette")

for (met_i in infer_methods) {
  for (met_m in merge_methods) {
    predict_clust(mtrx_dis, met_i, met_m)
  }
}
```

### Make clusters  
Now we are going to test hirechical clusters methods to asses for the clusters 


```{r}
merge_methods <- c()
res_coef <- c()
names(res_coef) <- merge_methods
for (clust_met in merge_methods) {
  #=====================================================================
  #hclust
  #=====================================================================
  #clust the distance object data with the method in the iteration
  clust_t <- hclust(dist_obj, method = clust_met)
  #save the Agglomerative coeficent
  res_coef[paste("HC_", clust_met, sep = "")] <- coef(clust_t)
  save_tree(clust_t, FALSE, paste("hclust_tree_", clust_met, sep = ""))
  #=====================================================================
  #agnes
  #=====================================================================
  clust_t <- tryCatch(
    {
      agnes(dist_obj, method = clust_met)
    },
    error = function(e) NULL
  )
  #save the Agglomerative coeficent
  tryCatch(
    {
      res_coef[paste("AG_", clust_met, sep = "")] <- clust_t$ac
      pltree(
        clust_t,
        cex = 0.6,
        hang = -1,
        main = paste("AG_", clust_met, sep = "")
      )
    },
    error = function(e) {
      res_coef[paste("AG_", clust_met, sep = "")] <- "None"
    }
  )
}
#=====================================================================
#diana
#=====================================================================
dClust <- diana(dist_obj)
pltree(dClust, cex = 0.6, hang = -1, main = "Diana Dendrogram")
res_coef["Diana"] <- dClust$dc

```